msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-06-03 paper.tex

In \msprime we have incorporated two examples of multiple-merger
coalescents.  One is a diploid extension \citep{BBE13} of the haploid
model of sweepstakes reproduction considered by \cite{EW06}, which is
a haploid Moran model adapted to sweepstakes reproduction.  Let $N$
denote population size, and take $\psi \in (0,1]$ to be fixed.  In
every generation, with probability $1-\varepsilon_N$ a single
individual (picked uniformly at random) perishes, and one of the
surviving individuals (sampled uniformly at random) produces one
offspring; with probability $\varepsilon_N$ a total of
$\lfloor \psi N \rfloor$ individuals perish, and of the remaining
individuals a single individual produces $\lfloor \psi N \rfloor -1 $
offspring.  Taking $\varepsilon_N = 1/N^\gamma$ for some $\gamma > 0$,
\cite{EW06} obtain specific examples of $\Lambda$-coalescents, where
the $\Lambda$ measure in Eq \eqref{lambdabk} is a point mass at
$\psi$.  The simplicity of this model does allow one to obtain some
explicit mathematical results (see e.g.\
\cite{EF2018,Matuszewski2017,Der2012,Freund2020}). The model
considered by \cite{EW06} has also been applied in algorithms for
simulating gene genealogies within phylogenies \citep{Zhu2015}. The
 specific model incorporated into \msprime is the diploid version
\citep{BBE13} of the model studied by \cite{EW06}, which would be
necessary in order to incorporate recombination.  In the model
considered by \cite{BBE13}, a single pair of diploid individuals
contribute offspring in each generation, selfing is excluded, and each
offspring is assigned one chromosome from each parent. There are,
therefore, four parent chromosomes involved in each reproduction
event, which can lead to up to four simultaneous mergers.  Let
\begin{esplit}{Cconst}
C_{b; k_1, \ldots, k_r;s } &=  \frac{4}{\psi^2} \sum_{\ell = 0}^{s \wedge (4-r)} \binom{s}{\ell} (4)_{r+\ell} (1-\psi)^{s-\ell }\left( \frac{\psi}{4} \right) ^{k_1 + \cdots + k_r + \ell},  \\
\end{esplit}
where  $C_{b; k_1, \ldots, k_r;s }$ corresponds to the coalescence rate $ \lambda_{b;k_1, \ldots, k_r;s}$ (see Eq. \eqref{xi}) of a  $\Xi$-coalescent on  $(\psi/4, \psi/4, \psi/4, \psi/4, 0, \ldots)$.
The total merger rate (see Eq.\ \eqref{lambdabkall}) is then 
\be\label{xidirlambdabk}
      \lambda(b;k_1, \ldots, k_r) =    \mathcal{N}(b; k_1, \ldots, k_r ) \left( \frac{c\psi^2/4}{1 +  c\psi^2/4}C_{b; k_1, \ldots, k_r;s } +     \frac{ \one{r=1, k_1 = 2} }{1 +  c\psi^2/4}  \right),
\ee
and the total coalescence rate (see Eq.\ \eqref{xilambdab}) becomes
    \begin{esplit}{xilambdab}
\lambda_b & =  \frac{c\psi^2/4}{1 +  c\psi^2/4}  \frac{4 }{\psi^2 } \left(1 - \sum_{\ell = 0 }^{b\wedge 4} \binom{b}{\ell} (4)_\ell \left( \frac{\psi}{4} \right)^\ell (1 - \psi)^{b-\ell } \right) +    \frac{1}{1 +  c\psi^2/4} \binom{b}{2}.  \\
\end{esplit}
The interpretation of Eq.\ \eqref{xilambdab} is that with probability  $1/(1 + c\psi^2/4)$  a `small' reproduction event occurs, in which the parent pair produce one  diploid offspring; with probability  $c\psi^2/4/(1 + c\psi^2/4)$ a `large' reproduction event occurs, in which  the parent pair produce  $\lfloor \psi N \rfloor$ offspring.

@@

\cite{Becheler2020}  investigate a general framework for  constructing efficient algorithms for general  coalescent processes.

>>>>

In \msprime we have incorporated two examples of multiple-merger
coalescents.  One is a diploid extension \citep{BBE13} of the haploid
model of sweepstakes reproduction considered by \cite{EW06}, which is
a haploid Moran model adapted to sweepstakes reproduction.  Let $N$
denote population size, and take $\psi \in (0,1]$ to be fixed.  In
every generation, with probability $1-\varepsilon_N$ a single
individual (picked uniformly at random) perishes, and one of the
surviving individuals (sampled uniformly at random) produces one
offspring; with probability $\varepsilon_N$ a total of
$\lfloor \psi N \rfloor$ individuals perish, and of the remaining
individuals a single individual produces $\lfloor \psi N \rfloor -1 $
offspring.  Taking $\varepsilon_N = 1/N^\gamma$ for some $\gamma > 0$,
\cite{EW06} obtain specific examples of $\Lambda$-coalescents, where
the $\Lambda$ measure in Eq \eqref{lambdabk} is a point mass at
$\psi$.  The simplicity of this model does allow one to obtain some
explicit mathematical results (see e.g.\
\cite{EF2018,Matuszewski2017,Der2012,Freund2020}). The model
considered by \cite{EW06} has also been applied in algorithms for
simulating gene genealogies within phylogenies \citep{zhu2015hybrid}. The
 specific model incorporated into \msprime is the diploid version
\citep{BBE13} of the model studied by \cite{EW06}, which would be
necessary in order to incorporate recombination.  In the model
considered by \cite{BBE13}, a single pair of diploid individuals
contribute offspring in each generation, selfing is excluded, and each
offspring is assigned one chromosome from each parent. There are,
therefore, four parent chromosomes involved in each reproduction
event, which can lead to up to four simultaneous mergers.  Let
\begin{esplit}{Cconst}
C_{b; k_1, \ldots, k_r;s } &=  \frac{4}{\psi^2} \sum_{\ell = 0}^{s \wedge (4-r)} \binom{s}{\ell} (4)_{r+\ell} (1-\psi)^{s-\ell }\left( \frac{\psi}{4} \right) ^{k_1 + \cdots + k_r + \ell},  \\
\end{esplit}
where  $C_{b; k_1, \ldots, k_r;s }$ corresponds to the coalescence rate $ \lambda_{b;k_1, \ldots, k_r;s}$ (see Eq. \eqref{xi}) of a  $\Xi$-coalescent on  $(\psi/4, \psi/4, \psi/4, \psi/4, 0, \ldots)$.
The total merger rate (see Eq.\ \eqref{lambdabkall}) is then
\be\label{xidirlambdabk}
      \lambda(b;k_1, \ldots, k_r) =    \mathcal{N}(b; k_1, \ldots, k_r ) \left( \frac{c\psi^2/4}{1 +  c\psi^2/4}C_{b; k_1, \ldots, k_r;s } +     \frac{ \one{r=1, k_1 = 2} }{1 +  c\psi^2/4}  \right),
\ee
and the total coalescence rate (see Eq.\ \eqref{xilambdab}) becomes
    \begin{esplit}{xilambdab}
\lambda_b & =  \frac{c\psi^2/4}{1 +  c\psi^2/4}  \frac{4 }{\psi^2 } \left(1 - \sum_{\ell = 0 }^{b\wedge 4} \binom{b}{\ell} (4)_\ell \left( \frac{\psi}{4} \right)^\ell (1 - \psi)^{b-\ell } \right) +    \frac{1}{1 +  c\psi^2/4} \binom{b}{2}.  \\
\end{esplit}
The interpretation of Eq.\ \eqref{xilambdab} is that with probability  $1/(1 + c\psi^2/4)$  a `small' reproduction event occurs, in which the parent pair produce one  diploid offspring; with probability  $c\psi^2/4/(1 + c\psi^2/4)$ a `large' reproduction event occurs, in which  the parent pair produce  $\lfloor \psi N \rfloor$ offspring.

@@

\cite{becheler2020occupancy}  investigate a general framework for  constructing efficient algorithms for general  coalescent processes.
